---
title: "Bayesian Supervised Learning in StochTree"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian-Supervised-Learning}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: vignettes.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to use the `bart()` function for Bayesian supervised learning 
(@chipman2010bart). To begin, we load the `stochtree` package.

```{r setup}
library(stochtree)
```

# Demo 1: Step Function

## Simulation

Here, we generate data from a simple step function.

```{r data}
# Generate the data
n <- 500
p_x <- 10
snr <- 3
X <- matrix(runif(n*p_x), ncol = p_x)
f_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (-7.5) + 
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (-2.5) + 
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2.5) + 
    ((0.75 <= X[,1]) & (1 > X[,1])) * (7.5)
)
noise_sd <- sd(f_XW) / snr
y <- f_XW + rnorm(n, 0, 1)*noise_sd

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
W_test <- NULL
W_train <- NULL
y_test <- y[test_inds]
y_train <- y[train_inds]
```

## Sampling and Analysis

### Warmstart

We first sample from an ensemble model of $y \mid X$ using "warm-start" 
initialization samples (@he2023stochastic). This is the default in 
`stochtree`.

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(sample_sigma_global = T, sample_sigma_leaf = T, num_trees_mean = 100)
bart_model_warmstart <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test, 
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the MCMC samples

```{r bart_warmstart_plot}
plot(bart_model_warmstart$sigma2_global_samples, ylab="sigma^2")
abline(h=noise_sd^2,col="red",lty=2,lwd=2.5)
plot(rowMeans(bart_model_warmstart$y_hat_test), y_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual")
abline(0,1,col="red",lty=2,lwd=2.5)
```

### BART MCMC without Warmstart

Next, we sample from this ensemble model without any warm-start initialization. 

```{r}
num_gfr <- 0
num_burnin <- 100
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(sample_sigma_global = T, sample_sigma_leaf = T, num_trees_mean = 100)
bart_model_root <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test, 
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc,
    params = bart_params
)
```

Inspect the MCMC samples

```{r bart_root_plot}
plot(bart_model_root$sigma2_global_samples, ylab="sigma^2")
abline(h=noise_sd^2,col="red",lty=2,lwd=2.5)
plot(rowMeans(bart_model_root$y_hat_test), y_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual")
abline(0,1,col="red",lty=2,lwd=2.5)
```

# Demo 2: Partitioned Linear Model

## Simulation

Here, we generate data from a simple partitioned linear model.

```{r data_plm}
# Generate the data
n <- 500
p_x <- 10
p_w <- 1
snr <- 3
X <- matrix(runif(n*p_x), ncol = p_x)
W <- matrix(runif(n*p_w), ncol = p_w)
f_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (-7.5*W[,1]) + 
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (-2.5*W[,1]) + 
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2.5*W[,1]) + 
    ((0.75 <= X[,1]) & (1 > X[,1])) * (7.5*W[,1])
)
noise_sd <- sd(f_XW) / snr
y <- f_XW + rnorm(n, 0, 1)*noise_sd

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
W_test <- W[test_inds,]
W_train <- W[train_inds,]
y_test <- y[test_inds]
y_train <- y[train_inds]
```

## Sampling and Analysis

### Warmstart

We first sample from an ensemble model of $y \mid X$ using "warm-start" 
initialization samples (@he2023stochastic). This is the default in 
`stochtree`.

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(sample_sigma_global = T, sample_sigma_leaf = T, num_trees_mean = 100)
bart_model_warmstart <- stochtree::bart(
    X_train = X_train, W_train = W_train, y_train = y_train, X_test = X_test, W_test = W_test, 
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the MCMC samples

```{r bart_warmstart_plot_plm}
plot(bart_model_warmstart$sigma2_global_samples, ylab="sigma^2")
abline(h=noise_sd^2,col="red",lty=2,lwd=2.5)
plot(rowMeans(bart_model_warmstart$y_hat_test), y_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual")
abline(0,1,col="red",lty=2,lwd=2.5)
```

### BART MCMC without Warmstart

Next, we sample from this ensemble model without any warm-start initialization. 

```{r}
num_gfr <- 0
num_burnin <- 100
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(sample_sigma_global = T, sample_sigma_leaf = T, num_trees_mean = 100)
bart_model_root <- stochtree::bart(
    X_train = X_train, W_train = W_train, y_train = y_train, X_test = X_test, W_test = W_test, 
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the BART samples after burnin.

```{r bart_root_plot_plm}
plot(bart_model_root$sigma2_global_samples, ylab="sigma^2")
abline(h=noise_sd^2,col="red",lty=2,lwd=2.5)
plot(rowMeans(bart_model_root$y_hat_test), y_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual")
abline(0,1,col="red",lty=2,lwd=2.5)
```

# Demo 3: Partitioned Linear Model with Random Effects

## Simulation

Here, we generate data from a simple partitioned linear model with an additive 
random effect structure.

```{r data_plm_rfx}
# Generate the data
n <- 500
p_x <- 10
p_w <- 1
snr <- 3
X <- matrix(runif(n*p_x), ncol = p_x)
W <- matrix(runif(n*p_w), ncol = p_w)
group_ids <- rep(c(1,2), n %/% 2)
rfx_coefs <- matrix(c(-5, -3, 5, 3), nrow=2, byrow=TRUE)
rfx_basis <- cbind(1, runif(n, -1, 1))
f_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (-7.5*W[,1]) + 
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (-2.5*W[,1]) + 
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2.5*W[,1]) + 
    ((0.75 <= X[,1]) & (1 > X[,1])) * (7.5*W[,1])
)
rfx_term <- rowSums(rfx_coefs[group_ids,] * rfx_basis)
noise_sd <- sd(f_XW) / snr
y <- f_XW + rfx_term + rnorm(n, 0, 1)*noise_sd

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
W_test <- W[test_inds,]
W_train <- W[train_inds,]
y_test <- y[test_inds]
y_train <- y[train_inds]
group_ids_test <- group_ids[test_inds]
group_ids_train <- group_ids[train_inds]
rfx_basis_test <- rfx_basis[test_inds,]
rfx_basis_train <- rfx_basis[train_inds,]
```

## Sampling and Analysis

### Warmstart

We first sample from an ensemble model of $y \mid X$ using "warm-start" 
initialization samples (@he2023stochastic). This is the default in 
`stochtree`.

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(sample_sigma_global = T, sample_sigma_leaf = T, num_trees_mean = 100)
bart_model_warmstart <- stochtree::bart(
    X_train = X_train, W_train = W_train, y_train = y_train, group_ids_train = group_ids_train, 
    rfx_basis_train = rfx_basis_train, X_test = X_test, W_test = W_test, group_ids_test = group_ids_test,
    rfx_basis_test = rfx_basis_test, num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the MCMC samples

```{r bart_warmstart_plot_plm_rfx}
plot(bart_model_warmstart$sigma2_global_samples, ylab="sigma^2")
abline(h=noise_sd^2,col="red",lty=2,lwd=2.5)
plot(rowMeans(bart_model_warmstart$y_hat_test), y_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual")
abline(0,1,col="red",lty=2,lwd=2.5)
```

### BART MCMC without Warmstart

Next, we sample from this ensemble model without any warm-start initialization. 

```{r}
num_gfr <- 0
num_burnin <- 100
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(sample_sigma_global = T, sample_sigma_leaf = T, num_trees_mean = 100)
bart_model_root <- stochtree::bart(
    X_train = X_train, W_train = W_train, y_train = y_train, group_ids_train = group_ids_train, 
    rfx_basis_train = rfx_basis_train, X_test = X_test, W_test = W_test, group_ids_test = group_ids_test,
    rfx_basis_test = rfx_basis_test, num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the MCMC samples

```{r bart_root_plot_plm_rfx}
plot(bart_model_root$sigma2_global_samples, ylab="sigma^2")
abline(h=noise_sd^2,col="red",lty=2,lwd=2.5)
plot(rowMeans(bart_model_root$y_hat_test), y_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual")
abline(0,1,col="red",lty=2,lwd=2.5)
```

# References
