---
title: "Bayesian Supervised Learning with Heteroskedasticity in StochTree"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Heteroskedasticity}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: vignettes.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to use the `bart()` function for Bayesian 
supervised learning (@chipman2010bart), with an additional "variance forest," 
for modeling conditional variance (see @murray2021log). To begin, we load the `stochtree` package.

```{r setup}
library(stochtree)
```

# Demo 1: Variance-Only Simulation (simple DGP)

## Simulation

Here, we generate data with a constant (zero) mean and a relatively simple covariate-modified variance function.

\begin{equation*}
\begin{aligned}
y &= 0 + \sigma^2(X) \epsilon\\
\sigma^2(X) &= \begin{cases}
0.25 & X_1 \geq 0 \text{ and } X_1 < 0.25\\
1 & X_1 \geq 0.25 \text{ and } X_1 < 0.5\\
4 & X_1 \geq 0.5 \text{ and } X_1 < 0.75\\
9 & X_1 \geq 0.75 \text{ and } X_1 < 1\\
\end{cases}\\
X_1,\dots,X_p &\sim \text{U}\left(0,1\right)\\
\epsilon &\sim \mathcal{N}\left(0,1\right)
\end{aligned}
\end{equation*}

```{r}
# Generate the data
n <- 500
p_x <- 10
X <- matrix(runif(n*p_x), ncol = p_x)
f_XW <- 0
s_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (0.5) + 
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (1) + 
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2) + 
    ((0.75 <= X[,1]) & (1 > X[,1])) * (3)
)
y <- f_XW + rnorm(n, 0, 1)*s_XW

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
W_test <- NULL
W_train <- NULL
y_test <- y[test_inds]
y_train <- y[train_inds]
f_x_test <- f_XW[test_inds]
f_x_train <- f_XW[train_inds]
s_x_test <- s_XW[test_inds]
s_x_train <- s_XW[train_inds]
```

## Sampling and Analysis

### Warmstart

We first sample the $\sigma^2(X)$ ensemble using "warm-start" 
initialization (@he2023stochastic). This is the default in 
`stochtree`.

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
num_trees <- 20
a_0 <- 1.5
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(num_trees_mean = 0, num_trees_variance = num_trees, 
                    sample_sigma_global = F, sample_sigma_leaf = F)
bart_model_warmstart <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test, 
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the MCMC samples

```{r}
plot(rowMeans(bart_model_warmstart$sigma_x_hat_test), s_x_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "standard deviation function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

### MCMC

We now sample the $\sigma^2(X)$ ensemble using MCMC with 
root initialization (as in @chipman2010bart).

```{r}
num_gfr <- 0
num_burnin <- 1000
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(num_trees_mean = 0, num_trees_variance = 50, 
                    sample_sigma_global = F, sample_sigma_leaf = F)
bart_model_mcmc <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test, 
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the MCMC samples

```{r}
plot(rowMeans(bart_model_mcmc$sigma_x_hat_test), s_x_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "standard deviation function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

# Demo 2: Variance-Only Simulation (complex DGP)

## Simulation

Here, we generate data with a constant (zero) mean and a more complex covariate-modified variance function.

\begin{equation*}
\begin{aligned}
y &= 0 + \sigma^2(X) \epsilon\\
\sigma^2(X) &= \begin{cases}
0.25X_3^2 & X_1 \geq 0 \text{ and } X_1 < 0.25\\
1X_3^2 & X_1 \geq 0.25 \text{ and } X_1 < 0.5\\
4X_3^2 & X_1 \geq 0.5 \text{ and } X_1 < 0.75\\
9X_3^2 & X_1 \geq 0.75 \text{ and } X_1 < 1\\
\end{cases}\\
X_1,\dots,X_p &\sim \text{U}\left(0,1\right)\\
\epsilon &\sim \mathcal{N}\left(0,1\right)
\end{aligned}
\end{equation*}

```{r}
# Generate the data
n <- 500
p_x <- 10
X <- matrix(runif(n*p_x), ncol = p_x)
f_XW <- 0
s_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (0.5*X[,3]) + 
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (1*X[,3]) + 
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2*X[,3]) + 
    ((0.75 <= X[,1]) & (1 > X[,1])) * (3*X[,3])
)
y <- f_XW + rnorm(n, 0, 1)*s_XW

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
W_test <- NULL
W_train <- NULL
y_test <- y[test_inds]
y_train <- y[train_inds]
f_x_test <- f_XW[test_inds]
f_x_train <- f_XW[train_inds]
s_x_test <- s_XW[test_inds]
s_x_train <- s_XW[train_inds]
```

## Sampling and Analysis

### Warmstart

We first sample the $\sigma^2(X)$ ensemble using "warm-start" 
initialization (@he2023stochastic). This is the default in 
`stochtree`.

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(num_trees_mean = 0, num_trees_variance = 50, 
                    alpha_mean = 0.95, beta_mean = 2, min_samples_leaf_mean = 5, 
                    alpha_variance = 0.95, beta_variance = 1.25, min_samples_leaf_variance = 1, 
                    sample_sigma_global = F, sample_sigma_leaf = F)
bart_model_warmstart <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test, 
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the MCMC samples

```{r}
plot(rowMeans(bart_model_warmstart$sigma_x_hat_test), s_x_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "standard deviation function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

### MCMC

We now sample the $\sigma^2(X)$ ensemble using MCMC with 
root initialization (as in @chipman2010bart).

```{r}
num_gfr <- 0
num_burnin <- 1000
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(num_trees_mean = 0, num_trees_variance = 50, 
                    alpha_mean = 0.95, beta_mean = 2, min_samples_leaf_mean = 5, 
                    alpha_variance = 0.95, beta_variance = 1.25, min_samples_leaf_variance = 5, 
                    sample_sigma_global = F, sample_sigma_leaf = F)
bart_model_mcmc <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test, 
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the MCMC samples

```{r}
plot(rowMeans(bart_model_mcmc$sigma_x_hat_test), s_x_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "standard deviation function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

# Demo 3: Mean and Variance Simulation (simple DGP)

## Simulation

Here, we generate data with (relatively simple) covariate-modified mean and variance functions.

\begin{equation*}
\begin{aligned}
y &= f(X) + \sigma^2(X) \epsilon\\
f(X) &= \begin{cases}
-6 & X_2 \geq 0 \text{ and } X_2 < 0.25\\
-2 & X_2 \geq 0.25 \text{ and } X_2 < 0.5\\
2 & X_2 \geq 0.5 \text{ and } X_2 < 0.75\\
6 & X_2 \geq 0.75 \text{ and } X_2 < 1\\
\end{cases}\\
\sigma^2(X) &= \begin{cases}
0.25 & X_1 \geq 0 \text{ and } X_1 < 0.25\\
1 & X_1 \geq 0.25 \text{ and } X_1 < 0.5\\
4 & X_1 \geq 0.5 \text{ and } X_1 < 0.75\\
9 & X_1 \geq 0.75 \text{ and } X_1 < 1\\
\end{cases}\\
X_1,\dots,X_p &\sim \text{U}\left(0,1\right)\\
\epsilon &\sim \mathcal{N}\left(0,1\right)
\end{aligned}
\end{equation*}

```{r}
# Generate the data
n <- 500
p_x <- 10
X <- matrix(runif(n*p_x), ncol = p_x)
f_XW <- (
    ((0 <= X[,2]) & (0.25 > X[,2])) * (-6) + 
    ((0.25 <= X[,2]) & (0.5 > X[,2])) * (-2) + 
    ((0.5 <= X[,2]) & (0.75 > X[,2])) * (2) + 
    ((0.75 <= X[,2]) & (1 > X[,2])) * (6)
)
s_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (0.5) + 
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (1) + 
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2) + 
    ((0.75 <= X[,1]) & (1 > X[,1])) * (3)
)
y <- f_XW + rnorm(n, 0, 1)*s_XW

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
W_test <- NULL
W_train <- NULL
y_test <- y[test_inds]
y_train <- y[train_inds]
f_x_test <- f_XW[test_inds]
f_x_train <- f_XW[train_inds]
s_x_test <- s_XW[test_inds]
s_x_train <- s_XW[train_inds]
```

## Sampling and Analysis

### Warmstart

We first sample the $\sigma^2(X)$ ensemble using "warm-start" 
initialization (@he2023stochastic). This is the default in 
`stochtree`.

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(num_trees_mean = 50, num_trees_variance = 50, 
                    alpha_mean = 0.95, beta_mean = 2, min_samples_leaf_mean = 5, 
                    alpha_variance = 0.95, beta_variance = 1.25, min_samples_leaf_variance = 5, 
                    sample_sigma_global = F, sample_sigma_leaf = F)
bart_model_warmstart <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test, 
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the MCMC samples

```{r}
plot(rowMeans(bart_model_warmstart$y_hat_test), y_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "mean function")
abline(0,1,col="red",lty=2,lwd=2.5)
plot(rowMeans(bart_model_warmstart$sigma_x_hat_test), s_x_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "standard deviation function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

### MCMC

We now sample the $\sigma^2(X)$ ensemble using MCMC with 
root initialization (as in @chipman2010bart).

```{r}
num_gfr <- 0
num_burnin <- 1000
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(num_trees_mean = 50, num_trees_variance = 50, 
                    alpha_mean = 0.95, beta_mean = 2, min_samples_leaf_mean = 5, 
                    alpha_variance = 0.95, beta_variance = 1.25, min_samples_leaf_variance = 5, 
                    sample_sigma_global = F, sample_sigma_leaf = F)
bart_model_mcmc <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test, 
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the MCMC samples

```{r}
plot(rowMeans(bart_model_mcmc$y_hat_test), y_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "mean function")
abline(0,1,col="red",lty=2,lwd=2.5)

plot(rowMeans(bart_model_mcmc$sigma_x_hat_test), s_x_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "standard deviation function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

# Demo 4: Mean and Variance Simulation (complex DGP)

## Simulation

Here, we generate data with more complex covariate-modified mean and variance functions.

\begin{equation*}
\begin{aligned}
y &= f(X) + \sigma^2(X) \epsilon\\
f(X) &= \begin{cases}
-6X_4 & X_2 \geq 0 \text{ and } X_2 < 0.25\\
-2X_4 & X_2 \geq 0.25 \text{ and } X_2 < 0.5\\
2X_4 & X_2 \geq 0.5 \text{ and } X_2 < 0.75\\
6X_4 & X_2 \geq 0.75 \text{ and } X_2 < 1\\
\end{cases}\\
\sigma^2(X) &= \begin{cases}
0.25X_3^2 & X_1 \geq 0 \text{ and } X_1 < 0.25\\
1X_3^2 & X_1 \geq 0.25 \text{ and } X_1 < 0.5\\
4X_3^2 & X_1 \geq 0.5 \text{ and } X_1 < 0.75\\
9X_3^2 & X_1 \geq 0.75 \text{ and } X_1 < 1\\
\end{cases}\\
X_1,\dots,X_p &\sim \text{U}\left(0,1\right)\\
\epsilon &\sim \mathcal{N}\left(0,1\right)
\end{aligned}
\end{equation*}

```{r}
# Generate the data
n <- 500
p_x <- 10
X <- matrix(runif(n*p_x), ncol = p_x)
f_XW <- (
    ((0 <= X[,2]) & (0.25 > X[,2])) * (-6*X[,4]) + 
    ((0.25 <= X[,2]) & (0.5 > X[,2])) * (-2*X[,4]) + 
    ((0.5 <= X[,2]) & (0.75 > X[,2])) * (2*X[,4]) + 
    ((0.75 <= X[,2]) & (1 > X[,2])) * (6*X[,4])
)
s_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (0.5*X[,3]) + 
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (1*X[,3]) + 
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2*X[,3]) + 
    ((0.75 <= X[,1]) & (1 > X[,1])) * (3*X[,3])
)
y <- f_XW + rnorm(n, 0, 1)*s_XW

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
W_test <- NULL
W_train <- NULL
y_test <- y[test_inds]
y_train <- y[train_inds]
f_x_test <- f_XW[test_inds]
f_x_train <- f_XW[train_inds]
s_x_test <- s_XW[test_inds]
s_x_train <- s_XW[train_inds]
```

## Sampling and Analysis

### Warmstart

We first sample the $\sigma^2(X)$ ensemble using "warm-start" 
initialization (@he2023stochastic). This is the default in 
`stochtree`.

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(num_trees_mean = 50, num_trees_variance = 50, 
                    alpha_mean = 0.95, beta_mean = 2, min_samples_leaf_mean = 5, 
                    alpha_variance = 0.95, beta_variance = 1.25, min_samples_leaf_variance = 5, 
                    sample_sigma_global = F, sample_sigma_leaf = F)
bart_model_warmstart <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test, 
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the MCMC samples

```{r}
plot(rowMeans(bart_model_warmstart$y_hat_test), y_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "mean function")
abline(0,1,col="red",lty=2,lwd=2.5)
plot(rowMeans(bart_model_warmstart$sigma_x_hat_test), s_x_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "standard deviation function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

### MCMC

We now sample the $\sigma^2(X)$ ensemble using MCMC with 
root initialization (as in @chipman2010bart).

```{r}
num_gfr <- 0
num_burnin <- 1000
num_mcmc <- 100
num_samples <- num_gfr + num_burnin + num_mcmc
bart_params <- list(num_trees_mean = 50, num_trees_variance = 50, 
                    alpha_mean = 0.95, beta_mean = 2, min_samples_leaf_mean = 5, 
                    alpha_variance = 0.95, beta_variance = 1.25, min_samples_leaf_variance = 5, 
                    sample_sigma_global = F, sample_sigma_leaf = F)
bart_model_mcmc <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test, 
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc, 
    params = bart_params
)
```

Inspect the MCMC samples

```{r}
plot(rowMeans(bart_model_mcmc$y_hat_test), y_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "mean function")
abline(0,1,col="red",lty=2,lwd=2.5)

plot(rowMeans(bart_model_mcmc$sigma_x_hat_test), s_x_test, 
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "standard deviation function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

# References
